Casual Bandits

### 2016-01-28 causal-bandit/icml-2015/general.tex

\newcommand{\calP}{\mathcal P}

Assume $Y \in \set{0,1}$. Let $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ be the
projection that selections the $N_i$ parents of variable $i$. Then we can write
\eq{
\P{X = x} = \prod_{k=1}^N \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
}
If we intervene by setting $X_i = j$ where $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$, then
\eq{
\P{X = x|do(X_i = j)} = \prod_{k \neq i} \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
}
Let $r(x) = \P{Y = 1|X = x}$. Then
\eq{
\P{Y = 1} = \sum_{x \in \set{0,1}^N} \P{X = x} r(x)
}
and
\eq{
\mu_{ij} 
& = \P{Y = 1|do(X_i = j)} \\
& = \sum_{x \in \set{0,1}^N : x_i = j} \P{X = x|do(X_i = j)} r(x)\,.
}
We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
intervening to set $X_i = j$. A natural choice is the importance weighted estimator
\eq{
Z_{ij} 
&= \frac{Y \ind{X_i = j}}{\P{X_i = j|\calP_i(X)}}\,. 
}
Checking:
\eq{
&\E{Z_{ij}} 
= \sum_x \frac{\P{X = x} r(x) \P{X = x | do(X_i = j)}}{\P{X = x}} \\
&= \P{Y = 1| do(X_i = j)} \,. 
}
The variance of $Z_{ij}$ may also be bounded
\eq{
&\Var[Z_{ij}]
\leq \E{Z_{ij}^2} \\
&\leq \sum_x \P{X = x} \left(\frac{\ind{X_i = j}}{\P{X_i = j|\calP_i(X) = \calP_i(x)}}\right)^2 \\
&= \E{\frac{1}{\P{X_i = j|\calP_i(X)}}} \\
&= V_{ij}\,.
}
In the general graph setting the dimension $m$ has a slightly different definition.
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}\,.
}

>>>>

\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}

We now consider single variable interventions on general causal graphs over variables $\boldsymbol{V}$. Let $Y \in \set{0,1}$ be the target variable and $\boldsymbol{X} = V/\set{Y \cup Desendants(Y)}$.

If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ from data collected without intervening via the truncated product formula \cite{Pearl2000} 1.37.

\eq {
P(Y,X) = \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\, \text{  $\forall \,\x$ consistent with $X_i = j$}  
}

Where $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ is the
projection that selections the $N_i$ parents of variable $i$.

We can then marginalize to obtain $P(Y|do(X_i = j))$

\eq {
P(Y|do(X_i = j)) =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\\
 =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\x}\\
 =& \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}\P{Y|\x}
}

Let $r(x) = \P{Y = 1|X = x}$. Then

\eq {
\mu_{ij} = P(Y = 1|do(X_i = j)) = \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}r(x)
}

We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
intervening to set $X_i = j$. A natural choice is the importance weighted estimator

\eq{
Z_{ij} 
&= \frac{Y \ind{X_i = j}}{\P{X_i = j|\calP_i(X)}}\,. 
}

Checking:
\eq{
&\E{Z_{ij}} 
= \sum_x \frac{\P{X = x} r(x) \P{X = x | do(X_i = j)}}{\P{X = x}} \\
&= \P{Y = 1| do(X_i = j)} \,. 
}
The variance of $Z_{ij}$ may also be bounded
\eq{
&\Var[Z_{ij}]
\leq \E{Z_{ij}^2} \\
&\leq \sum_x \P{X = x} \left(\frac{\ind{X_i = j}}{\P{X_i = j|\calP_i(X) = \calP_i(x)}}\right)^2 \\
&= \E{\frac{1}{\P{X_i = j|\calP_i(X)}}} \\
&= V_{ij}\,.
}
In the general graph setting the dimension $m$ has a slightly different definition.
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}\,.
}
